#### R Package installation and loading

Package checking and instalation

```{r}
if(!require(daltoolbox)) { 
  install.packages("daltoolbox")
}
```

Package loading

```{r}
library(daltoolbox)
```

#### Variable definitions and evaluation

Variable definition

```{r}
weight <- 60
height = 1.75
subject <- "A"
healthy <- TRUE
```

Variable evaluation

```{r}
weight
```

#### Functions for type checking

```{r}
is.numeric(weight) # variable 
is.double(weight)
is.integer(weight)
is.character(subject)
```

#### Functions for variable conversion

```{r}
weight <- as.integer(weight)
is.integer(weight)
```

#### Formulas and String formatting

Computing the body mass index (BMI) from the weight and height

```{r}
#Body mass index (BMI)
bmi <- weight/height^2 
bmi 
```

Functions for string manipulation

```{r}
message <- sprintf("%.1f", bmi)
print(message)
```

#### Vectors

definition

```{r}
weight <- c(60, 72, 57, 90, 95, 72) 
height <- c(1.75, 1.80, 1.65, 1.90, 1.74, 1.91)
subject <- c("A", "B", "C", "D", "E", "F")
```

evaluation

```{r}
weight
height
subject
```

#### Establishing the vector size

vector of size 10

```{r}
vec <- rep(0, 10)
vec
```

vector length

```{r}
length(weight)
```

#### Vector limits

from one to the length of the vector

```{r}
weight[1]
weight[length(weight)]
```

#### Iteration: for loop

from one to the length of weight

```{r}
bmi <- 0
for (i in 1:length(weight)) {
  bmi[i] <- weight[i]/height[i]^2
}
```

evaluation of the bmi vector

```{r}
bmi
```

#### Iteration: while loop

run while i is below or equal to the length of weight

```{r}
bmi <- 0
i <- 1
while (i <= length(weight)) {
  bmi[i] <- weight[i]/height[i]^2
  i <- i + 1
}
```

```{r}
bmi
```

#### Remove a variable

```{r}
rm(bmi)
exists("bmi")
```

#### Processing vectors at once


```{r}
bmi <- weight/height^2 
bmi 
```

#### Creating functions

name <- function(parameters) { body }

```{r}
compute_bmi <- function(weight, height) {
  bmi <- weight/height^2 
  return(bmi)
}
```

calling it

```{r}

bmi <- compute_bmi(60, 1.75)
bmi
```

#### Function processing vectors

```{r}
bmi <- compute_bmi(weight, height)
bmi
```

#### Example: function to compute the average

(iterating in all elements of the vector)

```{r}
average <- function(vec) {
    s <- 0
    n <- length(vec)
    for (x in vec) {
      s <- s + x  
    }
    return(s/n)
}
```

invoking the function

```{r}
avg_bmi <- average(bmi)
avg_bmi
```

#### Example of a function to compute the average (improved version)

manipulating vectors at once

```{r}
average <- function(vec) {
    s <- sum(vec)
    n <- length(vec)
    return(s/n)
}
```

invoking the function

```{r}
avg_bmi <- average(bmi)
avg_bmi
```

#### Example: function using mean function

Major statistical functions are available in R

```{r}
average <- function(vec) {
    return(mean(vec))
}
```

invoking the function

```{r}
avg_bmi <- average(bmi)
avg_bmi
```

#### Not available (NA)

Operations with NA lead to NA.

```{r}
x <- c(10, NA, 13)
y <- average(x)
y
```

Addressing NA with na.rm=TRUE

```{r}
average <- function(vec) {
    return(mean(vec, na.rm=TRUE))
}
```

```{r}
x <- c(10, NA, 13)
y <- average(x)
y
```

#### Plotting basic graphics

defining variables

```{r}
weight <- c(60, 72, 57, 90, 95, 72) 
height <- c(1.75, 1.80, 1.65, 1.90, 1.74, 1.91)
bmi <- weight/height^2 
```

scatter plots

```{r fig.width=6, fig.height=4.5}
plot(height, weight)
```

#### Most functions contain many default parameters

```{r fig.width=6, fig.height=4.5}
plot(height, weight, pch=2)
```

#### All functions in R that belongs to packages have help with examples

```{r}
?base::plot
```

#### Canvas for plotting is still active until a new plot

```{r fig.width=6, fig.height=4.5}
plot(height, weight)
hh = c(1.65, 1.70, 1.75, 1.80, 1.85, 1.90)
lines(hh, 22.5 * hh^2)
```

#### Factors

Factors are used to handle categorical data.

```{r}
pain <- c(0,3,2,2,1)
fpain <- factor(pain,levels=0:3, ordered=TRUE)
fpain
```

Levels provide correspondence between numerical values and categorical labels

```{r}
levels(fpain) <- c("none","mild","medium","severe")
fpain
```

#### Convert height to factor with levels: small, medium, high

coding setting element by element

```{r}
lev <- rep("", length(height))

for (i in 1:length(height)) {
  if (height[i] < 1.5)
    lev[i] <- "short"
  else if (height[i] < 1.9)
    lev[i] <- "medium"
  else 
    lev[i] <- "tall"
}

lev <- as.factor(lev)
lev
```

#### Using the function cut

```{r}
lev <- cut(height, breaks=c(0, 1.5, 1.9, .Machine$double.xmax), ordered=TRUE)
lev

levels(lev) <- c("short", "medium", "tall")
lev
```

#### Matrix

Matrices can be filled from vectors or data frames. 

```{r}
x <- 1:9
x
```

Converting a vector to matrix

```{r}
dim(x) <- c(3,3)
x
```

#### Scalar multiplication

```{r}
y <- 3*x
y
```

#### Matrix manipulation

Converting a vector to a matrix by row

```{r}
x <- matrix(1:9,nrow=3,byrow=T)
x
```

transposing a matrix

```{r}
x <- t(x)
x
```

#### Lists

Lists are used to work with "objects"

```{r}
a <- c(5260,5470,5640,6180,6390,6515,6805,7515,7515,8230,8770)
b <- c(3910,4220,3885,5160,5645,4680,5265,5975,6790,6900,7335)

mybag <- list(a, b, 0, "a")
mybag
```

#### Adding elements into a list

```{r}
n <- length(mybag)
mybag[[n+1]] <- "b"
mybag
```

#### List slicing

```{r}
slice <- mybag[1]
slice
is.list(slice)
```

#### Lists: member reference

Picking the first element

```{r}
h <- mybag[[1]]
h
```

An element can be evaluated. In this case, it is a vector.

```{r}
is.vector(h)
is.list(h)
```

#### Creating lists with attributes

They are properties on the list

```{r}
mybag <- list(x=a, y=b, const=0, lit="a")
mybag
```

#### Adding, accessing, and removing elements

```{r}
mybag$c <- mybag$x - mybag$y
mybag$const <- NULL
mybag$lit <- NULL
mybag
```

#### Data frames
Data frames (tables) provide support for structured data. 

```{r}
a <- c(5260,5470,5640,6180,6390,6515,6805,7515,7515,8230,8770)
b <- c(3910,4220,3885,5160,5645,4680,5265,5975,6790,6900,7335)
```

```{r}
data <- data.frame(A=a, B=b)
head(data)
```

#### Adding a column in a data frame 

```{r}
data$c <- data$A + data$B
head(data)
```

#### Removing a column of a data frame 

```{r}
data$A <- NULL
head(data)
```

####  Reading a csv file
There are many functions for reading CSV, Excel, and RData formats.

```{r}
wine = read.table(
    "http://archive.ics.uci.edu/ml/machine-learning-databases/wine/wine.data", 
                  header = TRUE, sep = ",")
  colnames(wine) <- c('Type', 'Alcohol', 'Malic', 'Ash', 
                      'Alcalinity', 'Magnesium', 'Phenols', 
                      'Flavanoids', 'Nonflavanoids',
                      'Proanthocyanins', 'Color', 'Hue', 
                      'Dilution', 'Proline')
head(wine)
```

####  saving in binary format 

saving a data frame


```{r}
save(wine, file="wine.RData")
```

removing a data frame from memory

```{r}
rm(wine)
```

loading it

```{r}
load("wine.RData")
head(wine, 3)
```

####  exporting data.frame into csv file

```{r}
write.table(wine, file="wine.csv", row.names=FALSE, quote = FALSE, sep = ",")
```

####  filtering vectors

```{r}
a <- c(5260,5470,5640,6180,6390,6515,6805,7515,7515,8230,8770)

# logical vector
bool <- (a > 7000)
bool
```

```{r}
# selecting elements from positions that are true
a[bool] 

a[a > 7000] 

```
####  filtering data frames

```{r}
a <- c(5260,5470,5640,6180,6390,6515,6805,7515,7515,8230,8770)
b <- c(3910,4220,3885,5160,5645,4680,5265,5975,6790,6900,7335)
data <- data.frame(a = a, b = b)
head(data[data$a > 7000,])
```

####  filtering data frames (rows and columns)

```{r}
head(data[data$a > 7000,c(1)])
```

#### Benchmarking performance of matrix and data frames

```{r}
rheight <- rnorm(100000, 1.8, sd=0.2)
rweight <- rnorm(100000, 72, sd=15)
```

#### Computing a column at once in a data frame

```{r}
start_time <- Sys.time()

hw <- data.frame(height=rheight, weight=rweight)
hw$bmi <- hw$weight/hw$height^2

end_time <- Sys.time()
end_time - start_time

object.size(hw)
```

#### Computing a column in a data frame cell by cell

```{r}
start_time <- Sys.time()
hw <- data.frame(height=rheight, weight=rweight)
for (i in 1:nrow(hw)) {
  hw$bmi[i] <- hw$weight[i]/hw$height[i]^2
}
end_time <- Sys.time()
end_time - start_time
```

#### Converting a data frame to a matrix, processing it, and going back

```{r}
start_time <- Sys.time()
hw <- data.frame(height=rheight, weight=rweight)

hw <- as.matrix(hw)

hw <- cbind(hw, 0)
for (i in 1:nrow(hw)) {
  hw[i,3] <- hw[i,2]/hw[i,1]^2
}

hw <- as.data.frame(hw)

end_time <- Sys.time()
end_time - start_time
```

#### Pipelines
The operator |> creates a pipeline. 

The first parameter of the next invoked function receives the data from the pipeline. 

Library $dplyr$ contains a set of functions that support relational algebra operations.

```{r}
flight_data <- read.table(text = "Year Quarter Flights Delays
                     2016 1 11 6
                     2016 2 12 5
                     2016 3 13 3
                     2016 4 12 5
                     2017 1 10 4
                     2017 2 9 3
                     2017 3 11 4
                     2017 4 25 15
                     2018 1 14 3
                     2018 2 12 5
                     2018 3 13 3
                     2018 4 15 4",
                     header = TRUE,sep = "")  
```

#### Displaying the data frame

```{r}
head(flight_data)
```

#### Basic Query


```{r}
library(dplyr)
```

```{r}
result <- flight_data |> 
   filter(Delays > 5) |> 
   select(Year, Quarter, Flights)
head(result)
```

#### Aggregated query


```{r}
result <- flight_data |> 
   group_by(Year) |> 
   summarize(mean = mean(Flights), sd = sd(Flights))
head(result)
```

#### Table join

Store table

```{r}
stores <- data.frame(
  city = c("Rio de Janeiro", "Sao Paulo", "Paris", "New York", "Tokyo"),
  value = c(10, 12, 20, 25, 18))
head(stores)
```

Division table

```{r}
divisions <- data.frame(
  city = c("Rio de Janeiro", "Sao Paulo", "Paris", "New York", "Tokyo"),
  country = c("Brazil", "Brazil", "France", "US", "Japan"))
head(divisions)
```

#### Merge function

The function $merge$ can be used to join data frames. It can be used to produce inner, left, right, and outer joins. 

```{r}
stdiv <- merge(stores, divisions, by.x="city", by.y="city")
head(stdiv)
```
#### Aggregating merged data frame

```{r}
result <- stdiv |> group_by(country) |> 
   summarize(count = n(), amount = sum(value))
head(result)
```

#### Statistical analysis
There are many statistical tests in R.
One of the most used is the t-test. It checks if the mean of observations is not different from a theoretical value.

```{r}
weight <- c(60, 72, 57, 90, 95, 72) 
height <- c(1.75, 1.80, 1.65, 1.90, 1.74, 1.91)
bmi <- weight/height^2 
```

```{r}
t.test(bmi, mu=22.5)
```

#### Python + R integration


Python code at retic.py
```text
import pyreadr
import pandas

def add(x, y):
  return x + y

def read_rdata_mem(data):
  x = data["x"]
  print(x)
  y = data["y"]
  data["z"] = x + y
  return(data)
```

Library reticulate enables seamless integration with Python
```{r}
library(reticulate)
source_python('https://raw.githubusercontent.com/eogasawara/analise-dados/refs/heads/main/python/retic.py')
x <- add(5, 10)
x
```


#### Python + R dataset integration

```{r}
data <- data.frame(x = c(1:5), y=c(11:15))
dfm <- read_rdata_mem(data)
head(dfm)
```
